\bf Approximating Likelihood Ratios with Calibrated Discriminative Classifiers

### 2016-01-21 jasa/paper-jasa.tex
Section 4: introduce parametrization of s

Thus far we have shown that the target likelihood ratio
$p(\mathbf{x}|\theta_0)/p(\mathbf{x}|\theta_1)$ with high dimensional features
$\mathbf{x}$ can be reproduced via the univariate densities
$p(s(\mathbf{x};\theta_0,\theta_1)|\theta_0)/p(s(\mathbf{x};\theta_0,\theta_1)|\theta_1)$
if the reduction $s(\mathbf{x}|\theta_0, \theta_1)$ is monotonic with
$p(\mathbf{x}|\theta_0)/p(\mathbf{x}|\theta_1)$. We now generalize from the
ratio of two simple hypotheses specified by $\theta_0$ and $\theta_1$ to the
case of composite hypothesis testing where $\theta$ are continuous model
parameters.
%
% In the case of a fixed classifier $s(x)$ it is possible to compute $s_e=s(x_e)$
% for the observed data and never refer back to the original features $x_e$. In
% the parametrized setting it is not possible to pre-compute $s(x_e; \theta_0,
% \theta_1)$ since $\theta_0$ and $\theta_1$ are unknown.
%
% The critical observation is that  if we postpone the evaluation of the
% classifier to the stage of evaluating the enveloping likelihood ratio, then we
% can identify the value of the parameters  that are being compared in the
% likelihood ratio with the values used as input to $s(x;\theta_0, \theta_1)$.
%
% \begin{equation}\label{eq:embedding}
% T(D; \theta_0, \theta_1) = \prod_e \frac{p(x_e|\theta_0)}{p(x_e|\theta_1)} = \prod_e  \frac{ p (s(x_e; \theta_0, \theta_1) | \theta_0)}
% {p (s(x_e; \theta_0,  \theta_1) | \theta_1) } \; .
% \end{equation}
%
% This is equivalent to approximating the likelihood function for $\theta_0$  when
% $\theta_1$ is held fixed.

@@

In the case of composite hypotheses $\theta \in \Theta_0$ against an alternative
$\theta \in \Theta_0^C$, the generalized likelihood ratio test, also known as
the profile likelihood ratio test, is commonly used
\begin{equation}\label{eqn:generalized-lr}
\lambda(\Theta_0) =  \frac{ \sup_{\theta \in \Theta_0} p({\cal D} | \theta)}{ \sup_{\theta \in \Theta} p({\cal D} | \theta)} \; .
\end{equation}
\glnote{For the denominator, shouldnt we have $\theta \in \Theta_0^C$ instead?}
This generalized likelihood ratio can be used both for hypothesis tests in the
presence of nuisance parameters or to create confidence intervals with or
without nuisance parameters.  Often, the parameter vector is broken into two
components $\theta=(\mu,\nu)$, where the $\mu$ components are considered
parameters of interest while the $\nu$ components are considered nuisance
parameters. In that case $\Theta_0$ corresponds to all values of $\nu$ with
$\mu$ fixed.

Let denote the maximum likelihood estimator
\begin{equation}\label{eq:mle}
\hat{\theta} = \argmax_\theta  p({\cal D} | \theta) \; .
\end{equation}
% and the conditional maximum likelihood estimator
% \begin{equation}\label{eq:cmle}
% \hat{\hat{\theta}} = \argmax_{\theta \in \Theta_0}  p({\cal D} | \theta) \; .
% \end{equation}
It is not obvious that if we are working with the distributions $p(s|\theta)$
(for some particular $s(\mathbf{x}; \theta_0, \theta_1)$ comparison) that we can find the
same estimators. Fortunately, there is a construction based on $p(s|\theta)$
that works. The maximum likelihood estimate of Eq.~\ref{eq:mle} is the same as
the value that maximizes the likelihood ratio with respect to $p({\cal
D}|\theta_1)$ for some fixed value of $\theta_1$. This allows us to use
Theorem~\ref{thm:ratio-equivalence} to reformulate the maximum likelihood estimate:
\begin{align}\label{eq:mle_withs}
\hat{\theta} &= \argmax_\theta  p({\cal D} | \theta) \nonumber \\
&= \argmax_\theta  \prod \frac{p(\mathbf{x}| \theta)}{p(\mathbf{x}|\theta_1)} \nonumber \\
&= \argmax_\theta  \prod \frac{p(s(\mathbf{x}; \theta, \theta_1) | \theta)}{p(s(\mathbf{x}; \theta, \theta_1) |\theta_1)} \; .
\end{align}
Note that it is important that we include the denominator $p(s(\mathbf{x}; \theta, \theta_1)
|\theta_1)$ because this cancels Jacobian factors that vary with $\theta$.

>>>>

Thus far we have shown that the target likelihood ratio
$r(\mathbf{x};\theta_0,\theta_1)=p(\mathbf{x}|\theta_0)/p(\mathbf{x}|\theta_1)$ with high dimensional features
$\mathbf{x}$ can be reproduced via the univariate densities
$p(s(\mathbf{x})|\theta_0)$ and $p(s(\mathbf{x})|\theta_1)$
if the reduction $s(\mathbf{x})$ is monotonic with
$r(\mathbf{x};\theta_0,\theta_1)$. We now generalize from the
ratio of two simple hypotheses specified by $\theta_0$ and $\theta_1$ to the
case of composite hypothesis testing where $\theta$ are continuous model
parameters.
%
% In the case of a fixed classifier $s(x)$ it is possible to compute $s_e=s(x_e)$
% for the observed data and never refer back to the original features $x_e$. In
% the parametrized setting it is not possible to pre-compute $s(x_e; \theta_0,
% \theta_1)$ since $\theta_0$ and $\theta_1$ are unknown.
%
% The critical observation is that  if we postpone the evaluation of the
% classifier to the stage of evaluating the enveloping likelihood ratio, then we
% can identify the value of the parameters  that are being compared in the
% likelihood ratio with the values used as input to $s(x;\theta_0, \theta_1)$.
%
% \begin{equation}\label{eq:embedding}
% T(D; \theta_0, \theta_1) = \prod_e \frac{p(x_e|\theta_0)}{p(x_e|\theta_1)} = \prod_e  \frac{ p (s(x_e; \theta_0, \theta_1) | \theta_0)}
% {p (s(x_e; \theta_0,  \theta_1) | \theta_1) } \; .
% \end{equation}
%
% This is equivalent to approximating the likelihood function for $\theta_0$  when
% $\theta_1$ is held fixed.

@@

In the case of composite hypotheses $\theta \in \Theta_0$ against an alternative
$\theta \in \Theta_0^C$, the generalized likelihood ratio test, also known as
the profile likelihood ratio test, is commonly used
\begin{equation}\label{eqn:generalized-lr}
\Lambda({\cal D}; \Theta_0) =  \frac{ \sup_{\theta \in \Theta_0} p({\cal D} | \theta)}{ \sup_{\theta \in \Theta} p({\cal D} | \theta)} \; .
\end{equation}
This generalized likelihood ratio can be used both for hypothesis tests in the
presence of nuisance parameters or to create confidence intervals with or
without nuisance parameters.  Often, the parameter vector is broken into two
components $\theta=(\mu,\nu)$, where the $\mu$ components are considered
parameters of interest while the $\nu$ components are considered nuisance
parameters. In that case $\Theta_0$ corresponds to all values of $\nu$ with
$\mu$ fixed.

Let denote the maximum likelihood estimator
\begin{equation}\label{eq:mle}
\hat{\theta} = \argmax_\theta  p({\cal D} | \theta) \; .
\end{equation}
% and the conditional maximum likelihood estimator
% \begin{equation}\label{eq:cmle}
% \hat{\hat{\theta}} = \argmax_{\theta \in \Theta_0}  p({\cal D} | \theta) \; .
% \end{equation}
It is not obvious that if we are working instead with distributions $p(s(\mathbf{x})|\theta)$
that we can find the
same estimators. Fortunately, there is a construction based on $s$
that works. The maximum likelihood estimate of Eqn.~\ref{eq:mle} is the same as
the value that maximizes the likelihood ratio with respect to $p({\cal
D}|\theta_1)$ for some fixed value of $\theta_1$. This allows us to use
Theorem~\ref{thm:ratio-equivalence} to reformulate the maximum likelihood estimate:
\begin{align}\label{eq:mle_withs}
\hat{\theta} &= \argmax_\theta  p({\cal D} | \theta) \nonumber \\
&= \argmax_\theta  \prod \frac{p(\mathbf{x}| \theta)}{p(\mathbf{x}|\theta_1)} \nonumber \\
&= \argmax_\theta  \prod \frac{p(s(\mathbf{x}; \theta, \theta_1) | \theta)}{p(s(\mathbf{x}; \theta, \theta_1) |\theta_1)} \;
\end{align}
where $s(\mathbf{x};\theta,\theta_1)$ denotes a \textit{parametrized} transformation $s$
of $X$ in terms of $(\theta,\theta_1)$. Note that it is important that we include the denominator $p(s(\mathbf{x}; \theta, \theta_1)
|\theta_1)$ because this cancels Jacobian factors that vary with $\theta$.
